---
title: "autocorrelation model"
author: "Kevin Kiley"
date: "6/2/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Autocorrelation Models

Loop that runs the lme model for all variables in the liwc data set. Takes quite a while to run. 

```{r}
nodupes <- liwc %>%
  group_by(hathi_author) %>% #10,417 cases
  filter(n() > 1) %>% #9,923 cases
  group_by(hathi_author, firstpub) %>% #some duplicates on first pub date
  filter(n() == 1) %>%
  arrange(hathi_author, firstpub) %>% #8,562 cases 
  group_by(hathi_author) %>%
  mutate(across(WC:filler, ~ .x - mean(.x)))
```

Models with year effects:

```{r}
m1 <- lme(affect ~ 1 + as.factor(firstpub), ~1 | hathi_author,
          data = nodupes)
BIC(m1) # 16991.19


m2 <- lme(affect ~ 1 + as.factor(firstpub), ~1 | hathi_author,
          correlation = corCAR1(form = ~ firstpub | hathi_author),
          data = nodupes)
BIC(m2) #17000.05

```

The model above includes year fixed-effects. In other words, it asks whether, above and beyond changes in the "mood" of authors that year, an author's work is more similar. Dropping the year effects but keeping the autocorrelation allows authors to follow trends and counts that as meaningful change. 

Models without year effects:

```{r}
m1.n <- lme(affect ~ 1, ~1 | hathi_author,
          data = nodupes)
BIC(m1.n) # 15614.86


m2.n <- lme(affect ~ 1, ~1 | hathi_author,
          correlation = corCAR1(form = ~ firstpub | hathi_author),
          data = nodupes)
BIC(m2.n) #15623.66

```


```{r}
d <- nodupes %>% select(hathi_author, firstpub, WC:filler) %>%
  pivot_longer(WC:filler, names_to = "variable", values_to = "y") %>% 
  arrange(variable) %>%
  drop_na() %>% 
  group_by(variable) %>% 
  nest()

```

```{r, message = FALSE, warning = FALSE}
#No autocorrelation, time effects
lme1_fit <- function (x) {
  lme(y ~ 1 + factor(firstpub), 
      random = ~1 | hathi_author,
      control = lmeControl(opt = "optim"),
      data = x)
}

#Autocorrelation, time effects
lme2_fit <- function(x) {
  lme(y ~ 1 + factor(firstpub), 
      random = ~1 | hathi_author,
      correlation = corCAR1(form = ~ firstpub | hathi_author),
      #control = lmeControl(opt = "optim"),
      data = x)
}

#No autocorrelation, no time effects
lme1_fit.n <- function (x) {
  lme(y ~ 1, 
      random = ~1 | hathi_author,
      control = lmeControl(opt = "optim"),
      data = x)
}

#Autocorrelation, no time effects
lme2_fit.n <- function(x) {
  lme(y ~ 1, 
      random = ~1 | hathi_author,
      correlation = corCAR1(form = ~ firstpub | hathi_author),
      #control = lmeControl(opt = "optim"),
      data = x)
}


lme_models <- d %>%
  mutate(lme1 = map(data, lme1_fit),
         lme2 = map(data, lme2_fit),
         lme1.n = map(data, lme1_fit.n),
         lme2.n = map(data, lme2_fit.n))

#save(lme_results, file = "~/Dropbox/data/lme_results.Rdata")

```



```{r}
results <- lme_models %>%
  pivot_longer(lme1:lme2.n, names_to = "mod_spec", values_to = "mod_object") %>% 
  mutate(glanced = map(mod_object, broom::glance)) %>% 
  unnest(glanced) %>% 
  select(variable, mod_spec, AIC, BIC, logLik) %>% 
  ungroup() %>%
  mutate(model = ifelse(mod_spec %in% c("lme1.n", "lme2.n"), "no year effect", "year effects"))


winners <- results %>%
  group_by(variable, model) %>% 
  mutate(BIC_diff = BIC - max(BIC)) %>% 
  filter(BIC == min(BIC)) %>% 
  mutate(type = mod_spec) %>%
  mutate(verdict = if_else(BIC_diff >= -2, "Inconclusive", type)) %>% 
  ungroup()

winners %>% left_join(liwc_categories, by = c("variable" = "liwc_var")) %>%
  filter(model == "year effects") %>%
   group_by(category, verdict) %>%
   summarise(n = n()) %>%
   group_by(category) %>%
   mutate(pct = n/sum(n)) %>%
  mutate(verdict = ifelse(verdict == "Inconclusive", "Inconclusive",
                          ifelse(verdict == "lme2", "AUM (RE + AR1)", "SDM (RE)"))) %>%
   ggplot(aes(x = category, y = pct, fill = verdict)) + 
   geom_bar(stat = "identity", color = "black") + 
   coord_flip() + 
   scale_fill_brewer(palette = "Set2") + 
   theme_minimal()


```


